\textit{Supporting Information for}: How is experience transformed into memory?

### 2018-09-05 paper/supporting_information.tex
paxton's and andy's edits to main text and supplement have been inputted

\title{\textit{Supporting information for}: How is experience transformed into memory?}
\author{Andrew C. Heusser, Paxton C. Fitzpatrick, and Jeremy R. Manning\\Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\Corresponding author: jeremy.r.manning@dartmouth.edu}

@@

\section*{Overview}
This document provides additional details about the methods we used in the main text.  We also include some additional figures referenced in the main text.

\section*{Additional details about topic modeling methods and results}
\subsection*{Optimizing topic model parameters}
In order to create accurate video and recall models, we used an optimization method that was driven by our ability to explain hand-annotated memory performance metrics collected by \cite{ChenEtal17}.  Specifically, we used a grid search to compute the $\omega$ (movie sliding window duration, in scenes), $\rho$ (recall sliding window duration, in sentences), and $K$ (number of topics) that satisfied
\[
\argmax_{\omega, \rho, K} \left[\mathrm{corr}\left(\mathrm{corr}\left(\mu\left(\omega, \rho, K\right), \nu\left(\omega, \rho, K\right)\right), \theta\right)\right],
\]
where $\mathrm{corr}(\mu, \nu)$ is the per-participant correlation between the upper triangles of the temporal correlation matrices of the video ($\mu$) and recall ($\nu$) trajectory, and $\theta$ is the per-participant hand-annotated memory performance.  We searched over a grid of pre-specified values for each of these parameters; the resulting correlations are displayed in Figure~\ref{fig:paramsearch}.  The optimal parameters were $\omega = 250$, $\rho = 50$, and $K = 100$.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/parameter_search}
\caption{\small \textbf{Optimizing topic model parameters.}  We performed a grid search over movie sliding window length ($\omega \in \{5, 10, 25, 50, 100 \}$), recall sliding window length ($\rho \in \{5, 10, 25, 50, 100 \}$, and number of topics ($K \in \{5, 10, 25, 50, 100 \}$.  The reported correlations are between per-subject video-recall trajectory correlations and per-subject hand-annotated memory performance ratings.}
\label{fig:paramsearch}
\end{figure}

@@

We also carried out an analysis of which annotated features tended to shape aspects of the video topic trajectory that were preserved in participants' recalls.  Specifically, we computed the timepoint-by-timepoint correlation matrix of the video topic trajectory, and correlated its upper triangle with the timepoint-by-timepoint correlation matrices of each participant's recall topic trajectory (resampled using linear interpolation to have the same number of timepoints as the video trajectory).  This yielded a single correlation coefficient for each participant.  We then carried out a series of analyses whereby we repeated the analysis with each annotated feature held out in turn.  Observing a lower correlation between the video and recall trajectories (when a given feature was held out) would indicate that the feature tends to be preserved in participants' recalls.  We found that hand-annotated narrative details were the most preserved type of feature, whereas information about the camera angle tended not to feature in participants' recalls (Fig.~\ref{fig:feature-importance}B).

Next, we wondered how the different types of features might relate.  For example, knowing which characters are on screen during a given scene may also provide information about which characters are speaking.  We computed video topic trajectories for each feature in turn, and then compared the temporal correlation matrices of all pairs of features.  This provided additional confirmation that the full model (including all types of features) was largely driven by narrative details.  We also found that character-driven features (characters on screen, characters speaking, and characters in focus) were strongly correlated.  Other details, such as the presense or absense of music, led to very different topic trajectories (Fig.~\ref{fig:feature-importance}C).

@@

\begin{figure}[tb]
\centering
\includegraphics[width=1\textwidth]{figs/list_learning}
\caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the video. \textbf{B}.  The probability of recalling each event, conditioned on having just recalled the event \textit{lag} events away in the video.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the video.  All panels: error bars denote bootstrap-estimated standard error of the mean.}
\label{fig:list-learning}
\end{figure}

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-video-events (34) matrix of zeros. Then for each participant, we found the index of the video event that was recalled first (i.e., the video event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 34 array representing the proportion of participant that recalled an event first, as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}A).

\paragraph{Lag conditional probability curve (lag-CRP).} The lag-CRP curve~\citep{Kaha96} reflects the probability of recalling a given event after the just-recalled event, as a function of their relative positions (or \textit{lag}).  In other words, a lag of 1 indicates that a recalled event came immediately after the previously recalled event in the video, and a lag of -3 indicates that a recalled event came 3 items before the previously recalled event.  For each recall transition (following the first recall), we computed the lag between the current recall event and the next recall event, normalizing by the total number of possible transitions.  This yielded a number-of-participants (17) by number-of-lags (-33 to +33; 67 lags total) matrix. We averaged over the rows of this matrix to obtain a group-averaged lag-CRP curve (Fig.~\ref{fig:list-learning}B).

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of their serial position during encoding. We initialized a number-of-participants (17) by number-of-video-events (34) matrix of zeros. Then, for each recalled event (and each participant), we found the index of the video event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix (i.e., for the given participant and event). This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero).  Finally, we averaged over the rows of the matrix to yield a 1 by 34 array representing the proportion of participants that recalled each event as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}C).

\paragraph{Temporal clustering scores.} Temporal clustering refers to the extent to which participants group their recall responses according to encoding position~\citep{PolyEtal09}. For instance, if a participant recalled the video events in the exact order they occurred, this would yield a score of 1.  If a participant recalled the video events in reverse order, this would yield a score of 0.  Similarly, if a participant recalled the events in random order, this yield yield an expected score of 0.5.  For each recall event transition (and separately for each participant), we sorted all not-yet-recalled events according to their absolute lag (i.e., distance away in the video).  We then computed the percentile rank of the next event the participant recalled.  We averaged these percentile ranks across all of the participant's recalls to obtain a single temporal clustering score for the participant (mean: 0.808, SEM: 0.022).  Overall, we found that participants with higher temporal clustering scores also tended to recall more events ($r = 0.62, p = 0.007$).

@@

\paragraph*{Precision.}
We tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the correlation between the topic vectors for each recall event and that of its closest-matching video event (only for the events which they recalled). We defined the precision as the average video-recall correlation across all of the events a participant recalled.  We found a strong correlation between hand-annotated memory performance and precision, suggesting that participants who remembered more events also remembered them more veridically ($r = 0.74, p = 0.0006$).

>>>>

\title{\textit{Supporting Information for}: How is experience transformed into memory?}
\author{Andrew C. Heusser, Paxton C. Fitzpatrick, and Jeremy R. Manning\\Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\Corresponding author: jeremy.r.manning@dartmouth.edu}

@@

\section*{Overview}
This document provides additional details about the methods we used in the main text.  We also include some additional analyses and figures referenced in the main text.

\section*{Additional details about topic modeling methods and results}
\subsection*{Optimizing topic model parameters}
In order to create accurate video and recall models, we used an optimization method that was driven by our ability to explain hand-annotated memory performance metrics collected by \cite{ChenEtal17}.  Specifically, we used a grid search to compute the $\omega$ (video sliding window duration, in scenes), $\rho$ (recall sliding window duration, in sentences), and $K$ (number of topics) that satisfied
\[
\argmax_{\omega, \rho, K} \left[\mathrm{corr}\left(\mathrm{corr}\left(\mu\left(\omega, \rho, K\right), \nu\left(\omega, \rho, K\right)\right), \theta\right)\right],
\]
where $\mathrm{corr}(\mu, \nu)$ is the per-participant correlation between the upper triangles of the temporal correlation matrices of the video ($\mu$) and recall ($\nu$) trajectory, and $\theta$ is the per-participant hand-annotated memory performance.  We searched over a grid of pre-specified values for each of these parameters; the resulting correlations are displayed in Figure~\ref{fig:paramsearch}.  The optimal parameters were $\omega = 50$, $\rho = 10$, and $K = 100$.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/parameter_search}
\caption{\small \textbf{Optimizing topic model parameters.}  We performed a grid search over video sliding window length ($\omega \in \{5, 10, 25, 50, 100 \}$), recall sliding window length ($\rho \in \{5, 10, 25, 50, 100 \}$, and number of topics ($K \in \{5, 10, 25, 50, 100 \}$.  The reported correlations are between per-subject video-recall trajectory correlations and per-subject hand-annotated memory performance ratings.}
\label{fig:paramsearch}
\end{figure}

@@

We also carried out an analysis of which annotated features tended to shape aspects of the video topic trajectory that were preserved in participants' recalls.  Specifically, we computed the timepoint-by-timepoint correlation matrix of the video topic trajectory, and correlated its upper triangle with that of the timepoint-by-timepoint correlation matrices of each participant's recall topic trajectory (resampled using linear interpolation to have the same number of timepoints as the video trajectory).  This yielded a single correlation coefficient for each participant.  We then repeated this analysis with each annotated feature held out in turn.  Observing a lower correlation between the video and recall trajectories (when a given feature was held out) would indicate that the feature tends to be preserved in participants' recalls.  We found that hand-annotated narrative details were the most preserved type of feature, whereas information about the camera angle tended not to influence participants' recalls (Fig.~\ref{fig:feature-importance}B).

Next, we wondered how the different types of features might relate.  For example, knowing which characters are on screen during a given scene may also provide information about which characters are speaking.  We computed video topic trajectories for each feature in turn, and then compared the temporal correlation matrices of all pairs of features.  This provided additional confirmation that the shape of the full trajectory (including all types of features) was largely driven by narrative details.  We also found that character-driven features (characters on screen, characters speaking, and characters in focus) were strongly correlated.  Other details, such as the presence or absence of music, led to very different topic trajectories (Fig.~\ref{fig:feature-importance}C).

@@

\begin{figure}[tb]
\centering
\includegraphics[width=1\textwidth]{figs/list_learning}
\caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the video. \textbf{B}.  The probability of recalling each event, conditioned on having most recently recalled the event \textit{lag} events away in the video.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the video.  All panels: error bars denote bootstrap-estimated standard error of the mean.}
\label{fig:list-learning}
\end{figure}

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-video-events (34) matrix of zeros. Then for each participant, we found the index of the video event that was recalled first (i.e., the video event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 34 array representing the proportion of participants that recalled an event first, as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}A).

\paragraph{Lag conditional probability curve (lag-CRP).} The lag-CRP curve~\citep{Kaha96} reflects the probability of recalling a given event after the just-recalled event, as a function of their relative positions (or \textit{lag}).  In other words, a lag of 1 indicates that a recalled event came immediately after the previously recalled event in the video, and a lag of -3 indicates that a recalled event came 3 events before the previously recalled event.  For each recall transition (following the first recall), we computed the lag between the current recall event and the next recall event, normalizing by the total number of possible transitions.  This yielded a number-of-participants (17) by number-of-lags (-33 to +33; 67 lags total) matrix. We averaged over the rows of this matrix to obtain a group-averaged lag-CRP curve (Fig.~\ref{fig:list-learning}B).

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of their serial position during encoding. We initialized a number-of-participants (17) by number-of-video-events (34) matrix of zeros. Then, for each recalled event, for each participant, we found the index of the video event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix (i.e., for the given participant and event). This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero).  Finally, we averaged over the rows of the matrix to yield a 1 by 34 array representing the proportion of participants that recalled each event as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}C).

\paragraph{Temporal clustering scores.} Temporal clustering refers to the extent to which participants group their recall responses according to encoding position~\citep{PolyEtal09}. For instance, if a participant recalled the video events in the exact order they occurred (or in exact reverse order), this would yield a score of 1.  If a participant recalled the events in random order, this would yield an expected score of 0.5.  For each recall event transition (and separately for each participant), we sorted all not-yet-recalled events according to their absolute lag (i.e., distance away in the video).  We then computed the percentile rank of the next event the participant recalled.  We averaged these percentile ranks across all of the participant's recalls to obtain a single temporal clustering score for the participant (mean: 0.808, SEM: 0.022).  Overall, we found that participants with higher temporal clustering scores also tended to recall more events ($r = 0.62, p = 0.007$).

@@

\paragraph*{Precision.}
We next tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the correlation between the topic vectors for each recall event and that of its closest-matching video event (only for the events which they recalled). We defined the precision as the average video-recall correlation across all of the events a participant recalled.  We found a strong correlation between hand-annotated memory performance and precision, suggesting that participants who remembered more events also remembered them more veridically ($r = 0.74, p = 0.0006$).
